
\documentclass[a4paper,onecolumn,11pt,unpublished]{quantumarticle}
\pdfoutput=1

%% Bibliography
\usepackage[numbers,sort&compress]{natbib}
\usepackage[nottoc]{tocbibind}

%% Math
\usepackage{amsmath,amssymb,amsthm,bm,amsfonts,mathrsfs,bbm}
\usepackage{braket}
\usepackage{physics}

%% Figures & tables
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{siunitx}

%% Algorithms & code
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{minted}     % requires --shell-escape; see README

%% Typography
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage[babel]{microtype}

%% Misc layout
\usepackage{wrapfig}
\usepackage{subcaption}

%% Theorem environments
\newtheorem{theorem}{Theorem}

%% Convenience macros
\newcommand{\QICK}{\textsc{qick}}
\newcommand{\todo}[1]{\textbf{\color{red}[TODO: #1]}}

\begin{document}

\title[Master Thesis -- Ender Jose Barillas]{%
  A General-Purpose QICK-Based Readout Architecture\\
  for Real-Time and Accelerated Quantum State Measurement}

\author{Ender Jose Barillas}
\supervisor{Joel \todo{Surname}}

\address{Departament de Física Quàntica i Astrofísica, Universitat de Barcelona \&
  Qilimanjaro Quantum Tech, 08001 Barcelona, Spain}
\ead{ender.barillas@qilimanjaro.tech}
\date{July 2026}

%% ----------------------------------------------------------------
\begin{abstract}
Low-latency and high-fidelity qubit state measurement is essential for
superconducting quantum computing, particularly for enabling mid-circuit
measurements and real-time feedback. Conventional readout techniques rely on
host-side processing, introducing delays that limit measurements to the end of a
quantum circuit. FPGA-based approaches offer a path toward in-situ state
discrimination, but many existing solutions are specialised or tightly coupled to
specific hardware designs.

This thesis presents a general-purpose readout architecture based on the
\QICK{} framework~\cite{Stefanazzi2022}, implemented on a Xilinx ZCU216 RFSoC
platform, that supports both mid-circuit and end-of-circuit quantum state
measurements. The work encompasses full system integration, including FPGA
configuration and remote operation, \QICK{}-based calibration, and the design of
a real-time readout data path. Control and readout waveforms are generated using
Direct Digital Synthesis (DDS), and the role of the analogue front end is
examined to ensure reliable signal transmission and measurement.

Building on this infrastructure, we investigate machine-learning-assisted state
discrimination within \QICK{} to accelerate readout while maintaining accuracy.
Lightweight machine-learning methods suitable for FPGA deployment are evaluated
with respect to latency and computational cost. The resulting architecture enables
low-latency state classification directly on the control hardware, reducing
reliance on host-side processing and providing a unified measurement pipeline for
dynamic quantum circuits.
\end{abstract}

\noindent{\it Keywords}: superconducting qubits, readout, FPGA, QICK, RFSoC,
  dispersive measurement, mid-circuit measurement, machine learning
\bigskip

%% ----------------------------------------------------------------
\section*{Acknowledgements}
I would like to thank the Quantronics team at Qilimanjaro Quantum Tech for their
guidance and support throughout this project.  Special thanks to Joel
\todo{Surname} for supervising this work and to all colleagues who offered
technical help and stimulating discussions.  This thesis was carried out during
an internship at Qilimanjaro from February to July 2026.

\bigskip
\tableofcontents
\clearpage

%% ----------------------------------------------------------------
\input{chapters/01_introduction}
\input{chapters/02_background}
\input{chapters/03_hardware}
\input{chapters/04_integration}
\input{chapters/05_readout}
\input{chapters/06_feedback}
\input{chapters/07_ml}
\input{chapters/08_conclusions}

%% ----------------------------------------------------------------
\bibliographystyle{unsrt}
\bibliography{references}

\clearpage
\appendix
\input{appendices/app_a_code}

\end{document}
