%% Chapter 2 — Theoretical Background
\section{Theoretical Background}
\label{sec:background}

\subsection{Superconducting Qubits}

Superconducting qubits are anharmonic LC oscillators whose nonlinearity is
provided by one or more Josephson junctions~\cite{Krantz2019}.  The Josephson
junction acts as a nonlinear inductor with energy $E_J\cos\hat{\varphi}$, where
$\hat{\varphi}$ is the superconducting phase across the junction and $E_J$ is the
Josephson energy.  Combined with a shunt capacitance $C$ (charging energy
$E_C = e^2/2C$), the system forms an artificial atom with a discrete, anharmonic
spectrum that can be addressed at the transition frequency $\omega_{01}$. Several qubit designs have been developed that exploit this anharmonicity in different parameter regimes; this thesis is primarily concerned with two: the transmon and the fluxonium.

\paragraph{Transmon.}
The transmon~\cite{Koch2007} operates deep in the regime $E_J \gg E_C$, making
its energy levels insensitive to charge noise.  Its transition frequency lies
typically in the \SIrange{4}{8}{\giga\hertz} range, and its anharmonicity
$\alpha = \omega_{12} - \omega_{01}$ is of order $-E_C/\hbar$, typically a few
hundred \si{\mega\hertz}.

\paragraph{Fluxonium.}
The fluxonium~\cite{Manucharyan2009} replaces the transmon's large capacitor with a
superinductance $L$ (inductive energy $E_L = (\Phi_0/2\pi)^2/L$, where $\Phi_0$ is
the flux quantum), threading the loop with an external flux $\Phi_z$.  The resulting
double-well potential can be tuned between a plasmon regime (near $\Phi_z = 0$) and
a fluxon regime (near $\Phi_z = \Phi_0/2$).  Fluxonium qubits offer long coherence
times and are of particular interest because their readout landscape is richer than
the transmon's: the dispersive shift $\chi$ varies strongly with flux bias, and near
the half-flux sweet spot the two eigenstates carry opposite persistent currents,
giving rise to a large state-dependent dispersive shift that can be exploited for
high-contrast readout~\cite{Manucharyan2009}.

\subsection{Circuit Quantum Electrodynamics}

In the circuit QED (cQED) architecture~\cite{Blais2004}, a qubit is coupled
dispersively to a microwave resonator.  The resonator acts both as a readout bus
and as a filter that protects the qubit from radiative decay into the measurement
chain.  The qubit--resonator Hamiltonian in the dipole coupling form
(setting $\hbar=1$) is
\begin{equation}
  \hat{H} = \hat{H}_\text{qubit}
           + \omega_r \hat{a}^\dagger \hat{a}
           + g\,\hat{n}_q (\hat{a} + \hat{a}^\dagger),
\label{eq:qubit_resonator}
\end{equation}
where $\omega_r$ is the resonator frequency, $\hat{a}$ ($\hat{a}^\dagger$) is the
resonator annihilation (creation) operator, $g$ is the qubit--resonator coupling
strength, and $\hat{n}_q$ is the relevant qubit operator (charge for capacitive
coupling)~\cite{Krantz2019}.  When the qubit--resonator detuning is large compared
to the coupling ($|g| \ll |\omega_{ij} - \omega_r|$), second-order perturbation
theory in $g$ yields an effective qubit-state-dependent shift of the resonator
frequency, derived in Section~\ref{subsec:dispersive} below.

\subsection{Dispersive Readout}
\label{subsec:dispersive}

When $|g_{ij}| \ll |\omega_{ij} - \omega_r|$ for all relevant qubit transitions
$i \leftrightarrow j$, second-order perturbation theory in $g$ yields a
qubit-state-dependent shift of the resonator frequency~\cite{Blais2021}:
\begin{equation}
  \hat{H}_\text{disp} = \left(\omega_r + \sum_i \chi_i \ket{i}\bra{i}\right)
  \hat{a}^\dagger \hat{a},
  \quad
  \chi_i = \sum_{j \neq i}
  \frac{|g_{ij}|^2}{\omega_{ij} - \omega_r}
  - \frac{|g_{ji}|^2}{\omega_{ji} - \omega_r},
  \label{eq:dispersive_shift}
\end{equation}
where $g_{ij} = \braket{i|g\,\hat{n}_q|j}$ and $\omega_{ij} = \omega_j -
  \omega_i$.  The textbook two-level dispersive shift $\chi = \chi_1 - \chi_0$
follows from truncating to the computational subspace, but for fluxonium-class
qubits the straddling regime (where $|g_{12}| \gg |g_{01}|$) makes this
truncation unsafe; higher levels must be included~\cite{Koch2007}.

In practice, one probes the resonator at its bare frequency $\omega_r$ and
observes a state-dependent phase shift of the transmitted or reflected signal.
For a single-qubit system, the resonator sits at $\omega_r + \chi_0$ when the
qubit is in $\ket{0}$ and at $\omega_r + \chi_1$ when in $\ket{1}$, resulting
in an IQ-plane separation that grows with the dimensionless cavity pull
$2|\chi|/\kappa$, where $\kappa$ is the resonator linewidth~\cite{Blais2004}.

\paragraph{Fluxonium dispersive readout.}
For a fluxonium qubit, the dispersive shift landscape is richer than for the
transmon because the qubit spectrum varies strongly with the external flux $\Phi_z$.
Near the half-flux point $\Phi_z = \Phi_0/2$, the double-well potential becomes
symmetric and the two lowest eigenstates carry opposite persistent currents.
The resulting state-dependent coupling to the resonator produces a dispersive shift
$\chi = \chi_1 - \chi_0$ whose magnitude depends strongly on the flux bias point,
making the readout contrast highly tunable.  This flux dependence is a key
motivation for the fluxonium readout optimisation work being pursued in related
projects within the Qilimanjaro team.

\subsection{Direct Digital Synthesis}
\label{subsec:dds}

Direct Digital Synthesis (DDS) generates analogue waveforms by accumulating a
phase register at a fixed clock rate and mapping the instantaneous phase value to
a digital amplitude via a look-up table.  A DDS numerically controlled oscillator
(NCO) produces a tone at frequency
\begin{equation}
  f_\text{out} = \frac{f_\text{clk} \cdot \Delta\phi}{2^N},
\end{equation}
where $f_\text{clk}$ is the system clock, $\Delta\phi$ is the phase increment per
clock cycle, and $N$ is the bit width of the phase accumulator.  DDS avoids the
spur and image problems associated with analogue IQ mixers because all
imperfections are digital and predictable.  In \QICK{}, the tProcessor drives DDS
engines on both the DAC (transmit) and ADC (receive) sides, enabling phase-locked
signal generation and coherent demodulation~\cite{Stefanazzi2022}.

\subsection{State Discrimination}
\label{subsec:state_discrim}

Following dispersive readout, the measured IQ quadrature amplitudes $(I, Q)$ form
two clusters in the complex plane — one for $\ket{0}$ and one for $\ket{1}$.
Single-shot state discrimination assigns each measurement to one of these
clusters.  The simplest approach is a linear threshold: a hyperplane in the IQ
plane that minimises the total misclassification probability, which for Gaussian
clusters of equal covariance is the perpendicular bisector of the line connecting
the centroids~\cite{Gambetta2007}.  The single-shot readout fidelity is
\begin{equation}
  \mathcal{F} = 1 - \frac{1}{2}\left(p(1|0) + p(0|1)\right),
\end{equation}
where $p(1|0)$ is the probability of assigning $\ket{1}$ given that the qubit was
prepared in $\ket{0}$, and vice versa.  More sophisticated classifiers — quadratic
discriminant analysis, neural networks — can improve fidelity when the clusters are
non-Gaussian or when multi-level leakage is present.

\subsection{Machine Learning for State Discrimination}
\label{subsec:ml_background}

The state discrimination problem defined in Section~\ref{subsec:state_discrim} is
a binary classification task: given a measurement outcome $\mathbf{x} \in
\mathbb{R}^d$, assign it to one of two classes ($\ket{0}$ or $\ket{1}$).  The
choice of classifier and its optimality depend on the statistical structure of the
data.

\paragraph{Linear Discriminant Analysis.}
If the two classes are distributed as multivariate Gaussians with equal covariance,
$p(\mathbf{x} \mid s) = \mathcal{N}(\boldsymbol{\mu}_s, \boldsymbol{\Sigma})$ for
$s \in \{0, 1\}$, then the Bayes-optimal decision boundary is the hyperplane
\begin{equation}
  (\boldsymbol{\mu}_1 - \boldsymbol{\mu}_0)^\top \boldsymbol{\Sigma}^{-1}
  \left(\mathbf{x} - \tfrac{1}{2}(\boldsymbol{\mu}_0 + \boldsymbol{\mu}_1)\right)
  = 0,
\label{eq:lda_boundary}
\end{equation}
which is exactly the decision boundary found by Linear Discriminant Analysis
(LDA)~\cite{Gambetta2007}.  Because this boundary is already optimal under the
Gaussian equal-covariance assumption, no more expressive classifier — including
multilayer perceptrons (MLPs) — can improve on it.  An MLP can at best approximate
this boundary; it cannot exceed the Bayes optimum.  When the covariances differ
between classes, Quadratic Discriminant Analysis (QDA) uses class-specific
covariances $\boldsymbol{\Sigma}_0, \boldsymbol{\Sigma}_1$ and yields a curved
boundary that is Bayes-optimal under the unequal-covariance Gaussian model.

\paragraph{Limitations of integrated IQ input.}
Both LDA and QDA operate on a single integrated $(I, Q)$ pair per shot — the
time-averaged quadrature amplitudes over the readout window.  This compression
discards all temporal information.  In particular, if the qubit relaxes from
$\ket{1}$ to $\ket{0}$ at some time $t^*$ within the readout window, the
integrator averages over both signal levels and produces a point that falls between
the two clusters.  No classifier operating on the integrated input can recover
these shots, regardless of its expressive power, because the information about the
true prepared state has been destroyed by the integration.

\paragraph{Convolutional Neural Networks on raw traces.}
A one-dimensional convolutional neural network (1D CNN) takes the full raw ADC
time trace as input, preserving the temporal structure that integration discards.
A CNN applies a bank of learned filters across the time axis via the convolution
operation
\begin{equation}
  (f * x)[t] = \sum_{\tau=0}^{k-1} f[\tau]\, x[t - \tau],
\label{eq:conv}
\end{equation}
where $f$ is a learned filter of length $k$ and $x$ is the input signal.  By
stacking convolutional layers, the network can detect local temporal features at
increasing scales — in the readout context, the step discontinuity in the IQ
trajectory that occurs when the qubit relaxes mid-shot.  This makes CNNs
well-suited to recovering $T_1$-relaxation shots that are irrecoverable from
integrated IQ, at the cost of requiring raw trace data rather than accumulated IQ
pairs.