%% Chapter 4 — System Integration
\section{System Integration}
\label{sec:integration}

\subsection{FPGA Configuration and Board Bring-up}

The ZCU216 boots the PYNQ Linux image from an SD card.  After boot, the
\QICK{} bitstream is loaded onto the FPGA fabric by instantiating a
\texttt{QickSoc} object in Python, which calls the PYNQ overlay loader and
configures all IP blocks.  Because the RFSoC converters are controlled through
AXI register maps rather than external chip-select lines, initialisation is
entirely software-driven.

The board is accessed over SSH.  To avoid re-initialising the hardware for every
new notebook session, a persistent Pyro4 object server is launched once on the
board and kept running:
\begin{minted}[fontsize=\small,linenos]{python}
# On the board (runs once per power cycle)
import Pyro4, qick
soc = qick.QickSoc()
daemon = Pyro4.Daemon()
uri = daemon.register(soc, "qick.soc")
Pyro4.locateNS().register("qick.soc", uri)
daemon.requestLoop()
\end{minted}

From a remote laptop:
\begin{minted}[fontsize=\small,linenos]{python}
import Pyro4
soc = Pyro4.Proxy("PYRONAME:qick.soc")
soccfg = soc  # lightweight config object
\end{minted}

This design preserves the phase calibration state (see
Section~\ref{subsec:phase_calib}) across reconnects and makes experiment
iteration fast.

\subsection{Channel Mapping}

The \QICK{} firmware assigns each physical DAC output to a numbered
\emph{generator channel} and each physical ADC input to a numbered
\emph{readout channel}.  On the ZCU216 with the default \QICK{} bitstream, the
mapping relevant to this work is shown in Table~\ref{tab:channels}.

\begin{table}[ht]
\centering
\caption{QICK channel mapping on the ZCU216.}
\label{tab:channels}
\begin{tabular}{llll}
\toprule
\QICK{} Channel & Physical Port & HD Header & Function \\
\midrule
Generator 6 & DAC 2\_231 & JHC3 & Readout / control drive \\
Readout 0   & ADC 0\_226 & JHC5 & IQ acquisition \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Phase Calibration}
\label{subsec:phase_calib}

Each time the ZCU216 is powered on or the FPGA bitstream is reloaded, the DDS
oscillators in the DAC and ADC blocks boot with a random mutual phase offset
$\phi$.  Without compensation, IQ measurements rotate unpredictably between
sessions, making it impossible to compare results across power cycles.

\paragraph{Phase model.}
A loopback tone from DAC 2\_231 to ADC 0\_226 measures $\phi$ at a given
frequency.  Because the two oscillators are effectively separated by a fixed time
delay $\tau$ (the total group delay around the loopback path), the phase
offset scales linearly with frequency:
\begin{equation}
  \phi(f) = \phi_0 - 360^\circ \cdot \tau \cdot f,
\label{eq:phase_model}
\end{equation}
where $\phi_0$ is a constant offset and $\tau$ is extracted from the fit.
Note that the fitted $\tau$ reflects the combined pipeline delay of the DAC
and ADC signal chains — including the digital up- and down-conversion stages
(DUC/DDC) and internal FIFOs — rather than the physical cable propagation delay,
which for a bench loopback of 1--2\,m contributes only a few nanoseconds.

\paragraph{Calibration procedure.}
The phase calibration sweeps a range of frequencies $f_i$ near the target readout
frequency, records the averaged IQ phasor at each point, and extracts
$\phi_i = \arg\langle I + jQ \rangle$.  The pair $(\phi_0, \tau)$ is then fitted
using \texttt{scipy.optimize.least\_squares}, taking care to handle the
$0^\circ/360^\circ$ phase wrap correctly.  The fit is considered reliable when
the signal-to-noise ratio satisfies $|\text{mag}| \gg \text{RMS}$ (e.g.
$637 \gg 1.8$~ADU in a typical calibration run).

The resulting \texttt{res\_phase} value:
\begin{equation}
  \phi_\text{res} = \phi_0 - 360^\circ \cdot \tau \cdot f_\text{res}
\end{equation}
is stored in the experiment configuration dictionary and passed to the readout
pulse via \texttt{soccfg.deg2reg(phi\_res)}, which converts degrees to the
integer register format expected by the tProcessor.  This calibration must be
re-run after every board reboot or bitstream reload, but remains valid throughout
a measurement session.

\paragraph{Measurement cadence.}
A summary of the recommended start-of-session checklist is:
\begin{enumerate}
  \item Power on and wait for PYNQ to boot ($\approx$\SI{60}{\second}).
  \item Connect the loopback (DAC 2\_231 $\to$ BALUN $\to$ SMA $\to$ BALUN $\to$
        ADC 0\_226) and run the phase calibration notebook.
  \item Identify the correct \texttt{adc\_trig\_offset} using decimated readout
        (see Section~\ref{subsec:decimated_vs_accum}).
  \item Switch to accumulated readout for all data-taking.
  \item At session end, call \texttt{soc.reset\_gens()} to halt all running
        generators.
\end{enumerate}
