%% Chapter 5 — Readout Architecture
\section{Readout Architecture}
\label{sec:readout}

\subsection{DDS-Based Signal Generation}

In a standard heterodyne readout chain, a local oscillator (LO) at frequency
$f_\text{LO}$ is mixed with a baseband envelope to up-convert the readout tone to
the resonator frequency $f_r$.  This analogue mixing process introduces image
products and LO leakage that degrade fidelity, requiring careful IQ imbalance
correction.

\QICK{} avoids analogue mixing entirely.  Both the readout drive and the
demodulation reference are produced by the on-chip DDS engines, which operate at
the direct output frequency of the RFSoC converters (up to several GHz).  The
waveform generated for qubit readout has the form
\begin{equation}
  s(t) = A \cdot \text{env}(t) \cdot \cos(2\pi f_r t + \phi_\text{res}),
\end{equation}
where $\text{env}(t)$ is the pulse envelope (constant, Gaussian, DRAG, etc.),
$f_r$ is programmed directly into the DDS frequency register, and
$\phi_\text{res}$ is the phase correction from calibration.  The only noise
sources are digital: quantisation, clock jitter, and DDS phase truncation spurs —
all of which are deterministic and can be mitigated by appropriate filtering and
sample-rate choice.

\subsection{Decimated vs. Accumulated Readout}
\label{subsec:decimated_vs_accum}

After the ADC samples the incoming signal, the \QICK{} readout block demodulates
it by multiplying with the NCO reference tone and low-pass filtering.  Two output
modes are available:

\paragraph{Decimated readout.}
Stores the full time-domain I/Q waveform over the readout window (e.g.\ 1000
samples).  This is essential for setup: it reveals when the readout pulse arrives
relative to the trigger (\texttt{adc\_trig\_offset}), whether the integration
window is correctly positioned, and whether the signal shape is as expected.  The
per-sample noise on the decimated output is $\sim$\SI{1.9}{ADU} RMS; this is the
raw sample-level noise floor, distinct from the per-shot integrated IQ noise
reported for accumulated readout below.

\paragraph{Accumulated readout.}
The FPGA sums all ADC samples within the readout window into a single $(I, Q)$
pair in hardware — one pair per shot.  This is the standard mode for SSRO: each
shot yields one IQ point, and the discrimination classifier operates on that
point.  The per-shot integrated IQ noise is determined by the within-window
averaging over $T_\text{win}$ samples, not by the number of repeated shots.
From the real SSRO data, the per-shot cluster widths are approximately
\SI{0.65}{}--\SI{0.97}{ADU} per quadrature (state-dependent), with an IQ
separation of \SI{2.53}{ADU} and a projected SNR of 2.46, yielding a single-shot
fidelity of 0.906~\cite{Walter2017}.

For applications that do \emph{not} require single-shot resolution — such as
resonator spectroscopy or Rabi calibration — the same shot can be repeated
$N_\text{reps}$ times and the results averaged, reducing the noise on the
\emph{mean} phasor as $1/\sqrt{N_\text{reps}}$:
\begin{equation}
  \sigma_\text{mean} = \frac{\sigma_\text{single}}{\sqrt{N_\text{reps}}}.
\end{equation}
This averaging is appropriate when only the average response is needed; it cannot
be applied to SSRO since averaging destroys the per-shot state information.

The practical workflow is: (1) use \texttt{acquire\_decimated()} to find the
correct \texttt{adc\_trig\_offset} and verify signal quality; (2) switch to
\texttt{acquire()} (accumulated) for all data-taking, using single shots for SSRO
and repeat-averaged shots for spectroscopy and calibration experiments.

\subsection{Resonator Spectroscopy}
\label{subsec:spectroscopy}

The resonator bare frequency $\omega_r$ and linewidth $\kappa$ are found by
sweeping the probe frequency around the estimated resonator frequency and measuring
the transmitted or reflected IQ amplitude.  In \QICK{}, there are two sweep
strategies:

\paragraph{Fast sweep (\texttt{RAveragerProgram}).}
The tProcessor increments the DAC DDS frequency register at each experiment step
via the \texttt{update()} method.  This approach is fast because the FPGA handles
all steps with no Python round-trips.  However, the ADC NCO frequency is set via
the ARM AXI bus and cannot be updated by the tProcessor.  For resonator
spectroscopy, both the DAC and ADC frequencies must track together; this is not
achievable with the fast method.

\paragraph{Slow sweep (Python outer loop).}
A \texttt{for} loop in Python updates both the DAC generator frequency and the
ADC readout frequency at each step.  The per-step overhead is a few milliseconds
(dominated by network round-trips in remote mode), but this approach can sweep
\emph{any} parameter, including the ADC NCO.  This is the correct approach for
resonator spectroscopy.

In practice:
\begin{minted}[fontsize=\small,linenos]{python}
freqs = np.linspace(f_start, f_stop, n_points)
amps  = []
for f in freqs:
    cfg["pulse_freq"] = f
    prog = AveragerProgram(soccfg, cfg)
    iq   = prog.acquire(soc, load_pulses=True, progress=False)
    amps.append(np.abs(iq[0][0][0] + 1j*iq[0][0][1]))
\end{minted}

The resonator frequency is identified as the dip (reflection) or peak
(transmission) in the swept amplitude response.

\subsection{Multiplexed Readout}
\label{subsec:multiplex}

For multi-qubit systems, simultaneous readout of several qubits on the same
physical line requires \emph{multiplexed readout}: a broadband probe signal
containing multiple tones at the individual resonator frequencies $\{f_{r,k}\}$~\cite{Heinsoo2018,Walter2017}.
The \QICK{} framework supports this through a \emph{Polyphase Filter Bank} (PFB)
readout mode.

A PFB is a bank of digital bandpass filters implemented as a single, efficient DSP
operation (polyphase decomposition of a prototype filter followed by an FFT).  It
takes the broadband ADC stream and simultaneously demodulates $N$ equally-spaced
frequency channels, analogous to a digital spectrum analyser.  Advantages over
the single-tone DDS approach include:
\begin{itemize}
  \item No need to retune the ADC NCO between qubits.
  \item Computational cost scales as $O(N \log N)$ rather than $O(N)$ independent
        demodulators.
  \item Sidesteps the ADC-frequency-sweep limitation: each PFB output channel is
        a fixed digital filter, not a tunable DDS.
\end{itemize}

Multiplexed readout via the \QICK{} PFB firmware was not implemented within the
scope of this work.  The single-tone DDS architecture described in
Sections~\ref{subsec:decimated_vs_accum} and~\ref{subsec:spectroscopy} was used
throughout.  A natural next step is to extend the architecture to cover the four
Dátil resonator frequencies characterised at Qilimanjaro: 6.505, 6.730, 6.978,
and \SI{7.389}{\giga\hertz}.  These frequencies lie above the \SIrange{5}{6}{\giga\hertz}
design passband of the XM655 BALUNs, so simultaneous multiplexed readout at all
four tones would require direct HD-header connections with discrete BALUNs, as
discussed in Section~\ref{sec:hardware}.  The PFB channel spacing and prototype
filter bandwidth would also need to be verified against the \SI{184}{\mega\hertz}
minimum separation between adjacent resonators in this device.
