%% Chapter 6 — Real-Time Feedback and Mid-Circuit Measurement
\section{Real-Time Feedback and Mid-Circuit Measurement}
\label{sec:feedback}

A key motivation for FPGA-based control is the ability to execute
\emph{real-time} conditional logic: perform a measurement, make a decision, and
apply a corrective pulse — all on the FPGA, without a round-trip to the host CPU.
This section describes the tProcessor-64 instruction set, demonstrates the
conditional branching primitive, and outlines the active qubit reset protocol.

\subsection{tProcessor Architecture}

The tProcessor-64 is a soft-processor synthesised in the FPGA fabric.  It
executes a custom instruction set tailored to pulse generation and readout
sequencing.  Key features include:
\begin{itemize}
  \item \textbf{Register file}: 32 general-purpose 64-bit registers per channel
        page, plus cross-page registers for inter-channel synchronisation.
  \item \textbf{Pulse triggers}: dedicated instructions that fire DAC signal
        generators or arm ADC acquisition windows.
  \item \textbf{Timing instructions}: \texttt{synci} (advance tProcessor time
        without touching channel timelines), \texttt{sync\_all} (synchronise all
        channels), \texttt{wait\_all} (stall until all readout windows close).
  \item \textbf{Conditional branch}: \texttt{condj page, reg\_a, op, reg\_b, label}
        jumps to \texttt{label} if \texttt{reg\_a op reg\_b} is true.
\end{itemize}

All branching targets are resolved at \emph{compile time} by the Python
assembler: Python's sequential execution of \texttt{body()} builds an
instruction list, \texttt{label()} records the current position as a named
address, and the assembler backpatches jump targets before loading the program
onto the tProcessor.  By the time the program runs, only integer addresses exist;
there is no runtime symbol table.

\subsection{Conditional Logic Demonstration}

The \QICK{} Demo 3 notebook demonstrates the core conditional-branch primitive:
\begin{minted}[fontsize=\small,linenos]{python}
def body(self):
    cfg = self.cfg
    self.trigger(adcs=[cfg["ro_ch"]],
                 adc_trig_offset=cfg["adc_trig_offset"])
    # condj: jump to SKIP_PULSE if number < threshold
    self.condj(0, 2, '<', 1, 'SKIP_PULSE')
    self.pulse(ch=cfg["res_ch"])          # skipped if condition true
    self.label('SKIP_PULSE')
    self.wait_all()
    self.sync_all(self.us2cycles(cfg["relax_delay"]))
\end{minted}

Registers 1 and 2 are pre-loaded with the threshold and the test value via
\texttt{regwi}.  When \texttt{number=100, threshold=50}: the condition
\texttt{100 < 50} is false, the pulse fires, and the ADC captures a loopback
signal.  When \texttt{number=10, threshold=50}: the condition is true, the pulse
is skipped, and the ADC trace is flat noise.

\paragraph{Timing subtlety.}
Inside a \texttt{body()} with an open ADC acquisition window, \texttt{synci}
must be used to advance time between pulses, not \texttt{sync\_all}.
\texttt{sync\_all} advances all channel timelines to the furthest point, which
is the end of the ADC window; subsequent pulses would then be scheduled after
the window has closed.  \texttt{synci} advances only the tProcessor's internal
counter, leaving the ADC timeline undisturbed.

\subsection{$\pi$ Pulse Calibration}
\label{subsec:pi_pulse}

A $\pi$ pulse rotates the qubit state by 180° on the Bloch sphere:
$\ket{0} \to \ket{1}$ and $\ket{1} \to \ket{0}$.  It is a fundamental primitive
for qubit initialisation (active reset) and single-qubit gates.  Calibrating it
requires a connected qubit chip; this section describes the procedure that would
be followed once the ZCU216 is connected to a device.

The $\pi$ pulse amplitude is found via a \emph{Rabi experiment}: the qubit drive
pulse amplitude is swept from zero, and the readout signal oscillates between the
$\ket{0}$ and $\ket{1}$ IQ clusters as the qubit is driven through successive
half-turns on the Bloch sphere.  The first amplitude at which the signal crosses
from the $\ket{0}$ cluster to the $\ket{1}$ cluster determines the $\pi$ pulse
amplitude.  The Rabi sequence is:
\begin{enumerate}
  \item Prepare qubit in $\ket{0}$ (wait $\gg T_1$).
  \item Apply qubit drive pulse at frequency $\omega_{01}$ with variable amplitude
        $A$ and fixed duration.
  \item Read out via the resonator.
\end{enumerate}

The qubit drive tone is at $\omega_{01}$ (the qubit transition frequency,
typically \SIrange{5}{7}{\giga\hertz} for fluxonium), which is distinct from the
resonator frequency $\omega_r$ used for readout.  Two separate generator channels
are therefore required: one for readout, one for qubit control.  The
\QICK{} firmware supports this natively — the tProcessor can address both channels
independently with nanosecond timing precision.

\subsection{Active Qubit Reset}
\label{subsec:active_reset}

Active reset rapidly returns the qubit to $\ket{0}$ by measuring its state and
applying a corrective $\pi$ pulse only if the qubit is found in $\ket{1}$.  The
\QICK{} tProcessor-64 has all the primitives required to implement this entirely
on-FPGA: the conditional branch demonstrated in Section~\ref{subsec:active_reset}
provides the decision logic, and the two-channel architecture supports
simultaneous readout and drive.  The following protocol was designed for
deployment once the board is connected to a qubit chip:
\begin{enumerate}
  \item \textbf{Dispersive readout}: fire the readout tone and acquire the
        accumulated IQ result.
  \item \textbf{Thresholding}: the readout firmware compares the IQ result against
        a pre-calibrated threshold and writes a binary decision (0 or 1) into a
        tProcessor register.
  \item \textbf{Conditional $\pi$ pulse}: \texttt{condj} checks the decision
        register.  If the qubit is in $\ket{1}$, a calibrated $\pi$ pulse is
        applied on the qubit drive channel.  Otherwise, the pulse is skipped.
  \item \textbf{Wait}: allow any residual photons in the resonator to decay before
        starting the experiment.
\end{enumerate}

Because the entire decision-to-pulse path would run inside the tProcessor, the
latency between completing the readout window and firing the corrective pulse is
determined solely by FPGA clock cycles.  For the ZCU216 operating at a
\SI{384}{\mega\hertz} fabric clock, this latency is of order tens of nanoseconds
— orders of magnitude faster than a software round-trip.

\subsection{Latency Budget}

Table~\ref{tab:latency} gives an estimated latency breakdown for the active reset
sequence.

\begin{table}[ht]
\centering
\caption{Estimated latency contributions in the active reset loop.}
\label{tab:latency}
\begin{tabular}{lrl}
\toprule
Stage & Latency & Notes \\
\midrule
Readout pulse duration     & \SIrange{0.5}{2}{\micro\second}   & Depends on $\kappa$ \\
ADC integration window     & \SIrange{0.5}{2}{\micro\second}   & \\
FPGA decision (threshold)  & $\sim$\SI{10}{\nano\second}        & tProcessor cycles \\
Conditional branch         & $\sim$\SI{3}{\nano\second}         & 1 tProc cycle \\
$\pi$ pulse duration       & \SIrange{50}{500}{\nano\second}    & Depends on $T_1$ \\
\midrule
Total (excl. readout)      & $<$\SI{1}{\micro\second}           & \\
\bottomrule
\end{tabular}
\end{table}

The dominant latency is the readout integration window; the digital decision and
conditional fire together contribute less than \SI{100}{\nano\second}, well within
qubit coherence times of \SIrange{10}{100}{\micro\second} for state-of-the-art
fluxonium devices.
