%% Chapter 7 — ML-Assisted State Discrimination
\section{ML-Assisted State Discrimination}
\label{sec:ml}

\subsection{IQ Data and the Discrimination Problem}

Each single-shot readout produces an averaged $(I, Q)$ pair — a point in the
complex plane.  Repeated preparation and measurement of the qubit in the same
state produces a cloud (cluster) of such points, whose centroid and spread depend
on the readout power, resonator parameters, and integration time.  Two such
clusters, one for $\ket{0}$ and one for $\ket{1}$, must be distinguished
reliably.

For ideal Gaussian clusters of equal covariance, the optimal linear classifier is
the perpendicular bisector of the line segment joining the two centroids:
\begin{equation}
  d(\mathbf{x}) = (\boldsymbol{\mu}_1 - \boldsymbol{\mu}_0)^\top
                  \boldsymbol{\Sigma}^{-1}
                  \left(\mathbf{x} - \tfrac{1}{2}(\boldsymbol{\mu}_0 + \boldsymbol{\mu}_1)\right)
  \underset{\hat{y}=0}{\overset{\hat{y}=1}{\gtrless}} 0,
\label{eq:lda}
\end{equation}
where $\boldsymbol{\mu}_0, \boldsymbol{\mu}_1 \in \mathbb{R}^2$ are the class
centroids and $\boldsymbol{\Sigma}$ is the common covariance matrix (Linear
Discriminant Analysis, LDA).  When the covariances differ, Quadratic Discriminant
Analysis (QDA) uses the class-specific covariances and yields a curved decision
boundary.

In practice, residual state preparation errors, qubit $T_1$ decay during the
readout pulse, and higher-level leakage ($\ket{2}$, $\ket{3}$) distort the
clusters from ideal Gaussians.  These effects motivate the use of more expressive
classifiers.  The assignment fidelity is defined as
\begin{equation}
  \mathcal{F} = \tfrac{1}{2}\bigl[P(\hat{s}=0\mid s=0) + P(\hat{s}=1\mid s=1)\bigr],
\label{eq:fidelity}
\end{equation}
the average per-class accuracy across both prepared states.

\subsection{Lightweight Classifiers for FPGA Deployment}
\label{subsec:classifiers}

For eventual on-FPGA deployment, the classifier must be computable using only
fixed-point arithmetic and a small number of multiply-accumulate operations.
The key constraints are:
\begin{itemize}
  \item \textbf{Latency}: the decision must complete within one tProcessor cycle
        budget ($\lesssim$\SI{10}{\nano\second}) to avoid extending the readout
        dead time.
  \item \textbf{Resource usage}: the classifier must fit in the available DSP
        slices and block RAM of the FPGA without displacing the readout and
        tProcessor logic.
  \item \textbf{Trainability}: the classifier must be trainable on a laptop from
        a few thousand calibration shots.
\end{itemize}

With these constraints in mind, three classifier families are considered.  A
\textbf{linear threshold} (LDA) reduces to a single dot product and a
comparison — the minimum possible arithmetic, directly implementable in the
tProcessor's \texttt{mathi} instruction.  A \textbf{quadratic threshold} (QDA)
evaluates a $2\times2$ matrix product (4 multiply-accumulates) and handles
asymmetric cluster covariances.  A \textbf{small MLP} with inputs $(I, Q)$, one
or two hidden layers of 4–64 neurons with ReLU activations, and a sigmoid output
can be computed as a matrix-vector product and fits in $\lesssim 200$ 18-bit DSP
slices on the ZCU216.  The following sections report results from evaluating
these classifiers offline on existing experimental data; FPGA deployment is
discussed in Section~\ref{subsec:deployment_path}.

\subsection{Experimental Data}
\label{subsec:ssro_data}

The dataset used throughout this chapter is \texttt{SSRO.h5}, collected on a
superconducting qubit device at Qilimanjaro.  It contains 2000 single-shot IQ
measurements per prepared state ($\ket{0}$ and $\ket{1}$), accumulated on-FPGA
by the rectangular integrator in \texttt{axis\_readout\_v2}.  Each shot is a
single $(I, Q)$ pair — the time-averaged quadrature amplitudes over the readout
window.  The class centroids and separations extracted from this dataset are
summarised in Table~\ref{tab:iq_stats}.

\begin{table}[ht]
\centering
\caption{IQ blob statistics extracted from \texttt{SSRO.h5}.}
\label{tab:iq_stats}
\begin{tabular}{lrr}
\toprule
Quantity & $\ket{0}$ & $\ket{1}$ \\
\midrule
Centroid $\mu_I$ (ADU)   & $-0.32$ & $-2.10$ \\
Centroid $\mu_Q$ (ADU)   & $+4.50$ & $+6.31$ \\
IQ separation (ADU)       & \multicolumn{2}{c}{$2.53$} \\
IQ correlation            & $-0.587$ & $-0.664$ \\
\bottomrule
\end{tabular}
\end{table}

The negative IQ correlation in both states indicates a rotated noise ellipse,
consistent with a readout tone that is not perfectly aligned to the I axis.  The
data were split 80/20 into training and test sets (3200 / 800 shots) using
stratified sampling.

\subsection{Part I — Integrated IQ Discrimination}
\label{subsec:integrated_iq}

LDA and a representative MLP (hidden layers 32$\to$16, ReLU) were trained on the
integrated IQ data.  Table~\ref{tab:integrated_results} reports the assignment
fidelity on the held-out test set.

\begin{table}[ht]
\centering
\caption{Assignment fidelity on integrated IQ data (test set, 800 shots).}
\label{tab:integrated_results}
\begin{tabular}{lrrr}
\toprule
Classifier & $\mathcal{F}$ & $\ket{0}$ acc. & $\ket{1}$ acc. \\
\midrule
LDA  & 0.906 & 0.940 & 0.873 \\
MLP (32$\to$16) & 0.906 & 0.940 & 0.873 \\
\bottomrule
\end{tabular}
\end{table}

LDA and the MLP achieve identical fidelity.  The $\ket{0}$ accuracy (94.0\%)
exceeds the $\ket{1}$ accuracy (87.3\%), pointing at a systematic source of
$\ket{1}$ misclassification.  This asymmetry is the first diagnostic signature of
$T_1$ relaxation during the readout window, which causes some $\ket{1}$ shots to
produce an integrated IQ that falls in the $\ket{0}$ blob.

\subsection{Part II — Architecture Search}
\label{subsec:arch_search}

To test whether the LDA–MLP parity in Part I is specific to the (32$\to$16)
architecture, a systematic grid search was performed over 20 MLP configurations
spanning 1–3 hidden layers and 4–64 neurons per layer (17 to 6465 parameters),
evaluated by 5-fold stratified cross-validation.

\begin{table}[ht]
\centering
\caption{Architecture grid search results (5-fold CV fidelity). LDA baseline:
         $0.898 \pm 0.013$.  No MLP architecture exceeds LDA.}
\label{tab:arch_search}
\begin{tabular}{lrrc}
\toprule
Architecture & $\mathcal{F}$ (CV) & $\pm$ std & $\Delta$ vs.\ LDA \\
\midrule
(4,)         & 0.897 & 0.012 & $-0.001$ \\
(8,)         & 0.897 & 0.012 & $-0.001$ \\
(16,)        & 0.898 & 0.012 & $-0.000$ \\
(32,)        & 0.898 & 0.012 & $-0.000$ \\
(64,)        & 0.897 & 0.012 & $-0.001$ \\
(8, 8)       & 0.896 & 0.012 & $-0.002$ \\
(16, 16)     & 0.897 & 0.012 & $-0.001$ \\
(32, 32)     & 0.897 & 0.012 & $-0.001$ \\
(64, 64)     & 0.897 & 0.012 & $-0.001$ \\
(16, 8)      & 0.897 & 0.012 & $-0.001$ \\
(32, 16)     & 0.897 & 0.012 & $-0.001$ \\
(16, 16, 8)  & 0.898 & 0.012 & $-0.000$ \\
(32, 32, 16) & 0.897 & 0.013 & $-0.001$ \\
(64, 64, 32) & 0.897 & 0.012 & $-0.002$ \\
\bottomrule
\end{tabular}
\end{table}

The result is striking: the fidelity landscape is entirely flat.  No architecture
— across three orders of magnitude in parameter count — improves on LDA.  This is
not a failure of the models; it is a statement about the data.

For two classes with Gaussian distributions $\mathcal{N}(\boldsymbol{\mu}_0,
\Sigma_0)$ and $\mathcal{N}(\boldsymbol{\mu}_1, \Sigma_1)$, the Bayes-optimal
boundary is a quadratic surface.  When $\Sigma_0 \approx \Sigma_1$ — as
approximately holds here, confirmed by the nearly identical IQ correlations in
Table~\ref{tab:iq_stats} — it collapses to the linear boundary found by LDA.  An
MLP can at best approximate this boundary; it cannot exceed the Bayes optimum.

\textbf{The implication is fundamental}: further fidelity improvement cannot come
from a more expressive classifier on the same integrated IQ input.  It must come
from richer input features.  The $\ket{1}$ accuracy deficit diagnosed in Part I
identifies the mechanism directly: $T_1$ relaxation mid-shot.

\subsection{Part III — Raw-Trace CNN}
\label{subsec:cnn}

\subsubsection{Motivation and the T1 Relaxation Problem}

When a $\ket{1}$ qubit relaxes to $\ket{0}$ at some time $t^*$ within the
readout window, the FPGA integrator averages over both signal levels and produces
an $(I, Q)$ point that lands between the two blobs.  Any classifier operating on
this compressed input — LDA, QDA, or MLP — must guess; the information about the
true prepared state has been discarded by the integration.

The raw ADC time trace, however, encodes $t^*$ directly as a step in the signal.
A temporal classifier trained on raw traces can in principle recover these shots
by detecting the jump.  Part III demonstrates this capability using
physically-grounded synthetic traces, since raw decimated traces from the
ZCU216/QICK setup are not yet available (acquiring them requires using
\texttt{acquire\_decimated} rather than the standard \texttt{acquire} call, and
collecting a dedicated \texttt{SSRO\_raw.h5} dataset with the board connected to
a qubit chip).

\subsubsection{Synthetic Trace Model}

Each shot is modelled as a noisy cavity response:
\begin{equation}
  \begin{pmatrix}i(t) \\ q(t)\end{pmatrix}
  = \frac{e(t)}{\bar{e}}\,\boldsymbol{\mu}_{s(t)} + \boldsymbol{\eta}(t),
  \qquad e(t) = 1 - e^{-t/\kappa},
\label{eq:trace_model}
\end{equation}
where $\bar{e} = \langle e(t)\rangle_t$ normalises the envelope so that
$\langle(i(t), q(t))\rangle_t = \boldsymbol{\mu}_{s(t)}$, matching the real blob
centroids exactly.  The noise $\boldsymbol{\eta}(t) \sim \mathcal{N}(\mathbf{0},
T\Sigma_s)$ is chosen so that the sample mean has covariance $\Sigma_s$, matching
the real blob widths.

For $\ket{1}$ shots, qubit relaxation is modelled by drawing a relaxation time
$t^*$ from a geometric distribution with rate $1 - e^{-1/T_1}$.  After $t^*$ the
signal switches from $\boldsymbol{\mu}_1$ to $\boldsymbol{\mu}_0$:
\begin{equation}
  \boldsymbol{\mu}_{s(t)} =
  \begin{cases}
    \boldsymbol{\mu}_1 & t < t^* \\
    \boldsymbol{\mu}_0 & t \geq t^*
  \end{cases}.
\end{equation}

The simulation parameters are: readout window $T = 200$ samples ($\approx
\SI{1}{\micro\second}$ at \SI{200}{MSps}), cavity decay $\kappa^{-1} = 20$
samples (\SI{100}{\nano\second}), and $T_1 = 600$ samples (\SI{3}{\micro\second}).
A self-consistency check confirmed that rectangular integration of the synthetic
traces reproduces the real \texttt{SSRO.h5} blob statistics to within 1\%
(Table~\ref{tab:self_consistency}).

\begin{table}[ht]
\centering
\caption{Self-consistency check: synthetic integrated IQ vs.\ real \texttt{SSRO.h5}.}
\label{tab:self_consistency}
\begin{tabular}{lrr}
\toprule
Quantity & Synthetic & Real \\
\midrule
$\mu_I$, $\ket{0}$            & $-0.341$ & $-0.324$ \\
$\mu_Q$, $\ket{0}$            & $+4.521$ & $+4.502$ \\
$\sigma_I$, $\ket{0}$         & $0.645$  & $0.649$ \\
$\mu_I$, $\ket{1}$ (no relax) & $-2.098$ & $-2.095$ \\
$\mu_Q$, $\ket{1}$ (no relax) & $+6.336$ & $+6.313$ \\
$\sigma_I$, $\ket{1}$ (no relax) & $0.851$ & $0.874$ \\
\bottomrule
\end{tabular}
\end{table}

With $T_1 = 600$ samples ($\approx 3\times$ the readout window), the expected
relaxation fraction among $\ket{1}$ shots is approximately 28--30\%, corresponding
to 566--603 relaxation events per 2000 shots.  This is notably higher than the
measured $\ket{1}$ error rate of approximately 12.75\% observed in
Table~\ref{tab:integrated_results}.  The discrepancy indicates that the simulated
$T_1$ is shorter than the actual qubit $T_1$ for this dataset, or equivalently
that $T_1$ should be set higher (around 1200--1500 samples) to match the observed
error rate.  The relative CNN-versus-LDA comparison remains valid regardless of
the exact $T_1$ value; the absolute fidelity numbers from Part III should be
treated as simulation estimates rather than predictions for the real system.

\subsubsection{CNN Architecture and Training}

The \texttt{ReadoutCNN} model operates on raw traces of shape $(T, 2)$.  Two
\texttt{Conv1d} layers compress the time axis, global average pooling collapses
it, and a small dense head produces the binary logit for $P(\ket{1})$.  The model
has 953 parameters, sized for deployment via \texttt{hls4ml}.

\begin{table}[ht]
\centering
\caption{\texttt{ReadoutCNN} architecture summary (953 parameters).}
\label{tab:cnn_arch}
\begin{tabular}{llr}
\toprule
Layer & Output shape & Parameters \\
\midrule
Conv1d(2$\to$8, k=5)  & $(T-4,\; 8)$  & 88  \\
ReLU + MaxPool1d(2)    & $((T-4)/2,\; 8)$ & — \\
Conv1d(8$\to$16, k=5) & $(\cdot,\; 16)$ & 656 \\
ReLU + GlobalAvgPool  & $(16,)$       & — \\
Linear(16$\to$8)      & $(8,)$        & 136 \\
ReLU + Linear(8$\to$1)& $(1,)$        & 73  \\
\bottomrule
\end{tabular}
\end{table}

The model was trained for 60 epochs using binary cross-entropy loss and the Adam
optimiser.  Test fidelity reached 0.994 by epoch 39 and remained stable
thereafter.

\subsubsection{Results}

Table~\ref{tab:cnn_results} compares LDA on integrated IQ against the CNN on raw
traces, evaluated on the same held-out test set.

\begin{table}[ht]
\centering
\caption{CNN vs.\ LDA on synthetic raw traces.  LDA operates on the
         time-averaged $(I,Q)$; the CNN operates on the full trace $(T,2)$.
         All values are from the held-out test set.}
\label{tab:cnn_results}
\begin{tabular}{lrrrr}
\toprule
Classifier & $\mathcal{F}$ & $\ket{0}$ acc. & $\ket{1}$ acc. (all) &
             $\ket{1}$ acc. (relax.\ only) \\
\midrule
LDA (integrated IQ)  & 0.826 & 0.875 & 0.778 & 0.443 \\
1D CNN (raw trace)   & 0.994 & 0.995 & 0.993 & 0.984 \\
\midrule
CNN gain             & $+16.8\%$ & $+12.0\%$ & $+21.5\%$ & $+54.1\%$ \\
\bottomrule
\end{tabular}
\end{table}

The CNN achieves 98.4\% accuracy on relaxation shots, compared to 44.3\% for LDA
— a gain of 54 percentage points on the shots that are entirely invisible to
integration-based methods.  The $P(\ket{1})$ confidence as a function of
relaxation time $t^*$ rises monotonically: shots that relax late in the window
(spending most of the trace in $\ket{1}$) are classified with near-certainty,
while shots that relax early (spending most of the trace in $\ket{0}$) are
correctly identified as ambiguous.  This confirms the CNN is exploiting the
temporal jump structure, not merely interpolating between blob positions.

These results should be understood as simulation estimates.  The absolute fidelity
numbers depend on the synthetic $T_1$ value (Section~\ref{subsec:cnn}), which does
not precisely match the real qubit's $T_1$.  The qualitative conclusion — that a
CNN on raw traces can recover $T_1$-relaxation shots that are irrecoverable from
integrated IQ — is robust to this uncertainty.

\subsection{FPGA Deployment Path}
\label{subsec:deployment_path}

FPGA deployment of the trained classifiers was not achieved within the scope of
this work.  This section outlines the concrete steps required to close the loop
from the trained models to real-time discrimination inside the QICK firmware.

The current readout pipeline — rectangular integrator followed by the tProcessor
threshold register — already implements LDA implicitly.  Formalising this as an
explicit dot-product instruction (\texttt{mathi}) adds no hardware overhead.  The
CNN requires a separate data path: raw decimated traces from
\texttt{acquire\_decimated} are fed into an \texttt{hls4ml}-synthesised IP block
that produces a binary decision within approximately \SI{100}{\nano\second} to
\SI{200}{\nano\second} at \SI{250}{\mega\hertz}, well within the tProcessor's
feedback budget (Table~\ref{tab:latency}).

The full deployment sequence is:
\begin{enumerate}
  \item \textbf{Collect real raw traces}: connect the ZCU216 to the qubit chip and
        acquire \texttt{SSRO\_raw.h5} using \texttt{acquire\_decimated} (shape
        $(2, N_\text{shots}, T, 2)$).  Set \texttt{readout\_length} to at least
        $3/\kappa$ to capture the full cavity ringdown.
  \item \textbf{Retrain and validate}: retrain the CNN on real traces.  Recalibrate
        the synthetic $T_1$ parameter to match the measured $\ket{1}$ error rate
        ($\approx 12.75\%$), which requires $T_1 \approx 1200$--$1500$ samples.
  \item \textbf{Export to HLS}: export the trained PyTorch model to ONNX, then
        convert using \texttt{hls4ml} targeting the ZCU216 fabric
        (\texttt{xczu49dr-ffvf1760-2-e}), \texttt{io\_stream}, and fixed-point
        precision \texttt{ap\_fixed<16,6>}.
  \item \textbf{Synthesise and integrate}: synthesise the HLS project in Vivado
        and integrate the resulting IP block into the QICK bitstream alongside the
        existing readout chain.
  \item \textbf{Benchmark}: compare LDA-threshold, CNN-IP, and host-side
        discrimination on the same experimental data to quantify latency reduction
        and fidelity gain under realistic qubit conditions.
\end{enumerate}

Based on the model's 953 parameters and typical \texttt{hls4ml} resource scaling for comparable architectures, the CNN IP block is expected to require on the order of \num{10000} LUTs and tens of DSP slices — a small fraction of the ZCU216's available fabric — though this must be confirmed by synthesis.
