%% Chapter 8 — Conclusions
\section{Conclusions and Outlook}
\label{sec:conclusions}

\subsection{Summary}

This thesis has documented the setup, characterisation, and initial evaluation of
a \QICK{}-based readout architecture for superconducting qubits on a Xilinx
ZCU216 RFSoC platform at Qilimanjaro Quantum Tech, and has explored the design of
a machine-learning-assisted state discrimination pipeline using existing
experimental data.

The main contributions are:

\begin{enumerate}
  \item \textbf{System integration}: The ZCU216 board, XM655 analogue front-end,
        and \QICK{} firmware were brought up from scratch, including SSH access,
        PYNQ configuration, and a persistent Pyro4 remote-operation server that
        preserves calibration state across notebook sessions.

  \item \textbf{Phase-coherent readout calibration}: A systematic calibration
        procedure was developed that measures the random DAC–ADC DDS phase offset
        at board power-on.  The linear phase-vs-frequency model
        $\phi(f) = \phi_0 - 360^\circ \tau f$ was validated, and the fitted
        \texttt{res\_phase} value was incorporated into the standard experiment
        configuration.

  \item \textbf{Readout data-path}: The trade-offs between decimated (waveform)
        and accumulated (averaged IQ) readout modes were characterised.  The
        accumulated mode produces one integrated $(I, Q)$ pair per shot, suitable
        for SSRO; repeat-averaging across shots improves spectroscopy SNR but
        destroys single-shot information and cannot be used for state
        discrimination.  The resonator spectroscopy workflow was established for
        both single-qubit and multiplexed scenarios.

  \item \textbf{Real-time feedback primitives}: The tProcessor V2 conditional
        branch (\texttt{condj}) was demonstrated in loopback as the core primitive
        for mid-circuit measurement.  An active qubit reset sequence was designed,
        combining dispersive readout, thresholding, and a calibrated $\pi$ pulse,
        with an estimated digital decision-to-fire latency below
        \SI{100}{\nano\second}.

  \item \textbf{ML-assisted discrimination}: A complete offline discrimination
      pipeline was developed and evaluated on existing experimental SSRO data.
      On integrated IQ data, LDA achieves an assignment fidelity of 90.6\%
      ($\ket{0}$: 94.0\%, $\ket{1}$: 87.3\%), and a systematic grid search over
      20 MLP architectures confirms that no nonlinear classifier improves on
      this — a consequence of the near-Gaussian, linearly separable IQ blob
      structure.  The $\ket{1}$ accuracy deficit is identified as the signature
      of $T_1$ relaxation mid-shot, which integration-based methods cannot
      recover.  A lightweight 1D CNN (953 parameters) trained on
      physically-grounded synthetic raw traces achieves 99.9\% overall fidelity
      and 100\% accuracy on the 91 relaxation shots in the test set, compared
      to 41.8\% for LDA on the same shots — a gain of 58.2 percentage points.
      The 100\% figure reflects near-certain classification on the available
      test shots and should be interpreted in light of the small relaxation
      sample size.  FPGA deployment of the CNN was not achieved within the
      scope of this work; Section~\ref{subsec:deployment_path} outlines the
      concrete path toward it.
\end{enumerate}

\subsection{Outlook}

Several natural extensions of this work follow directly from the results above:

\paragraph{CNN retraining on real raw traces.}
The CNN results in Chapter~\ref{sec:ml} are based on synthetic traces anchored to
real IQ statistics.  The first experimental priority is to collect a real
\texttt{SSRO\_raw.h5} dataset from the ZCU216 using \texttt{acquire\_decimated},
retrain the CNN, and validate that the $T_1$-relaxation recovery generalises to
real data.  The synthetic $T_1$ parameter should be recalibrated to match the
measured $\ket{1}$ error rate ($\approx 12.75\%$, implying $T_1 \approx
1200$--$1500$ samples rather than the 600 used in simulation) before the
synthetic-to-real transfer is attempted.

\paragraph{On-FPGA classifier deployment.}
With real raw traces in hand, the CNN can be exported via \texttt{hls4ml} and
synthesised as a fixed-point IP block in the QICK bitstream, as described in
Section~\ref{subsec:deployment_path}.  This closes the active-reset feedback loop
at full speed: the tProcessor receives a binary decision from the CNN IP block
within \SI{100}{\nano\second}--\SI{200}{\nano\second}, without a software
round-trip.

\paragraph{Multi-qubit readout.}
The current architecture supports a single readout tone.  Extending to
multiplexed readout via the \QICK{} PFB firmware would allow simultaneous readout
of all qubits in Qilimanjaro's multi-qubit chips, which is a prerequisite for
running quantum error-correction cycles.

\paragraph{Integration with Qilimanjaro's software stack.}
The \QICK{}-based readout layer developed here should be integrated with
Qilimanjaro's higher-level quantum programming interfaces (circuit compilation,
calibration databases, experiment management) to provide a production-ready
control stack.

\paragraph{Fluxonium-specific readout optimisation.}
The dispersive shift landscape for fluxonium qubits is two-dimensional in the
$(\Phi_z, \Phi_x)$ flux space.  An ML-assisted optimisation of the readout
circuit parameters — coupling capacitance, resonator frequency — to maximise the
dimensionless cavity pull over a large flux region is a complementary research
direction being pursued in a related BSc project within the team.
